---
title: "Counterfactual Surfaces: Shooter development versus Shot-selection deployment"
subtitle: "Figure source for *Modern NBA Shot Shot selection*"
author: "Shane Sanders"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(dplyr)
library(ggplot2)
```

## Why not ggplot2

`ggplot2` has no 3D geometry and no plans for one; its grammar is built on 2D positional
aesthetics. Three routes give a genuine surface in R, in increasing order of control:

1. **`plot3D::persp3D`** — direct perspective surfaces with a colour ramp, closest to the figure
   reproduced here. Used below.
2. **`rayshader`** — extrudes a `ggplot2` object into 3D and raytraces it. Handsome, slow, and it
   maps the *fill* aesthetic to height, so the surface must first be written as a raster geom.
3. **`plotly::plot_ly(type = "surface")`** — interactive and the easiest to read on screen, but
   produces HTML rather than vector output for a paper.

A 2D alternative that loses nothing for this particular story is a filled contour of the frontier
with the actual path overlaid, which `ggplot2` handles natively. That version is given last.

## Data

```{r data}
G <- read.csv("cf_surface_grid.csv")   # season, risk, z_dev_frozen, z_sel_frozen
P <- read.csv("cf_paths.csv")          # season, sS/mS, sD/mD, sA/mA
str(G)
```

`z_dev_frozen` is the frontier with conversion rates held at 2013-14 in every season;
`z_sel_frozen` lets conversion evolve as observed. The paths are the league's realised portfolio
under each counterfactual: `sS, mS` vary shot selection only, `sD, mD` vary shooter development only.

## Surfaces via plot3D

```{r persp, eval=FALSE}
library(plot3D)
yrs  <- sort(unique(G$season)); rsk <- sort(unique(G$risk))
ZA   <- matrix(G$z_dev_frozen, nrow = length(yrs), byrow = TRUE)
ZB   <- matrix(G$z_sel_frozen, nrow = length(yrs), byrow = TRUE)

panel <- function(Z, path_x, path_y, path_z, pal, ttl) {
  persp3D(x = yrs, y = rsk, z = Z, col = pal(100), colkey = FALSE,
          theta = -62, phi = 20, expand = 0.6, border = NA, shade = 0.25,
          alpha = 0.62, zlim = c(103, 123), bty = "b2",
          xlab = "season", ylab = "risk: scoring SD", zlab = "points per game",
          main = ttl, cex.main = 0.95)
  lines3D(path_x, path_y, path_z, col = "white", lwd = 6, add = TRUE)
  lines3D(path_x, path_y, path_z, col = pal(100)[85], lwd = 4, add = TRUE)
  points3D(path_x, path_y, path_z, col = "white", pch = 21, cex = 0.8, add = TRUE)
}
par(mfrow = c(1, 2), mar = c(2, 2, 3, 1))
panel(ZA, P$season, P$sS, P$mS, colorRampPalette(c("#FFF7BC", "#C0392B")),
      "(a) Shooter development frozen, shot selection as observed")
panel(ZB, P$season, P$sD, P$mD, colorRampPalette(c("#F7FCB9", "#1B7A4B")),
      "(b) Shot selection frozen, shooter development as observed")
```

## The ggplot2-native alternative

```{r gg, fig.width=11, fig.height=4.2}
long <- G %>%
  tidyr::pivot_longer(c(z_dev_frozen, z_sel_frozen),
                      names_to = "panel", values_to = "frontier") %>%
  mutate(panel = factor(panel,
           levels = c("z_dev_frozen", "z_sel_frozen"),
           labels = c("Shooter development frozen, shot selection as observed",
                      "Shot selection frozen, shooter development as observed")))

paths <- bind_rows(
  transmute(P, season, risk = sS, pts = mS,
            panel = "Shooter development frozen, shot selection as observed"),
  transmute(P, season, risk = sD, pts = mD,
            panel = "Shot selection frozen, shooter development as observed"))

ggplot(long, aes(season, risk)) +
  geom_raster(aes(fill = frontier), interpolate = TRUE) +
  geom_contour(aes(z = frontier), colour = "white", linewidth = 0.25, bins = 12) +
  geom_path(data = paths, aes(season, risk), colour = "white", linewidth = 2.1) +
  geom_path(data = paths, aes(season, risk), colour = "grey15", linewidth = 1.1) +
  geom_point(data = paths, aes(season, risk), colour = "white",
             fill = "grey15", shape = 21, size = 1.7, stroke = 0.6) +
  facet_wrap(~ panel) +
  scale_fill_viridis_c(option = "inferno", name = "frontier\n(points)") +
  labs(x = NULL, y = "risk: scoring standard deviation",
       title = "The frontier and the actual path under each counterfactual",
       subtitle = paste("Left: the frontier is flat and the actual path moves right.",
                        "Right: the frontier rises and the actual path rises with it.")) +
  theme_minimal(base_size = 11) +
  theme(panel.grid = element_blank(), strip.text = element_text(face = "bold"),
        plot.title = element_text(face = "bold"), plot.title.position = "plot")
```

## Reading

With shooter development frozen, the frontier rises by 0.0 points and the actual path rises by
`r sprintf("%+.2f", P$mS[nrow(P)] - P$mS[1])` points while moving right:
`r sprintf("%+.2f", P$sS[nrow(P)] - P$sS[1])` points of risk for
`r sprintf("%+.2f", P$mS[nrow(P)] - P$mS[1])` points of scoring. With shot selection frozen, the frontier rises by
`r sprintf("%.1f", max(G$z_sel_frozen) - max(G$z_dev_frozen))` points and the actual path rises by `r sprintf("%+.2f", P$mD[nrow(P)] - P$mD[1])` points for
`r sprintf("%+.2f", P$sD[nrow(P)] - P$sD[1])`. Together the two changes deliver
`r sprintf("%+.2f", P$mA[nrow(P)] - P$mA[1])` points against
`r sprintf("%+.2f", (P$mS[nrow(P)]-P$mS[1]) + (P$mD[nrow(P)]-P$mD[1]))` if they were additive: an
interaction of
`r sprintf("%+.2f", (P$mA[nrow(P)]-P$mA[1]) - ((P$mS[nrow(P)]-P$mS[1]) + (P$mD[nrow(P)]-P$mD[1])))`.
