---
title: "Risk and Return in NBA Team-Game Shot-Selection Portfolios"
subtitle: "Full reproduction of the MIT SSAC27 abstract"
author: "Shane Sanders · Department of Sport Analytics, Falk College of Sport, Syracuse University"
output:
  pdf_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center", dpi = 200)
library(lpSolve)
library(ggplot2)
library(dplyr)
library(tidyr)
dir.create("derived", showWarnings = FALSE)
dir.create("figures", showWarnings = FALSE)
```

# What this reproduces

Each NBA shot is a Bernoulli trial scaled by its point value, so a team's shot
distribution is a portfolio with both a return and a risk. Attempts are independent
trials rather than correlated holdings, so covariance vanishes, portfolio variance is
linear in the attempt-share weights, and the efficient frontier solves a linear rather
than a quadratic program.

This document recomputes every number reported in the abstract from
`team_season_zones.csv` alone, checks each against the published value, and redraws both
figures.

# Constants and data

```{r constants}
SHARE <- c("s03", "s310", "s1016", "s163", "sc3", "snc3")
CONV  <- c("p03", "p310", "p1016", "p163", "pc3", "pnc3")
NAME  <- c("0-3 ft", "3-10 ft", "10-16 ft", "16 ft-arc", "corner 3", "above break 3")
PTS   <- c(2, 2, 2, 2, 3, 3)

# shot-pursuant free-throw value per attempt, by era and shot type, recovered from the
# box-score accounting identity Phi = (1 - w) theta_2 + w theta_3
theta_of <- function(season) {
  tt <- if (season <= 2017) c(0.207, 0.216)
        else if (season <= 2021) c(0.196, 0.202)
        else c(0.250, 0.124)
  ifelse(PTS == 2, tt[1], tt[2])
}
games_of <- function(season) if (season == 2020) 71 else if (season == 2021) 72 else 82

Z <- read.csv("team_season_zones.csv")
stopifnot(all(c("season", "ab", "q", "FGA", SHARE, CONV) %in% names(Z)))
seasons <- sort(unique(Z$season))
cat(sprintf("%d team-seasons, %d seasons: %d to %d\n",
            nrow(Z), length(seasons), min(seasons), max(seasons)))
```

```{r helpers}
# league-representative portfolio for one season
league <- function(season) {
  g <- Z[Z$season == season, ]
  S <- as.matrix(g[, SHARE]); S <- S / rowSums(S)
  list(x = colMeans(S),
       p = colMeans(as.matrix(g[, CONV])),
       q = mean(g$q),
       n = mean(g$FGA))
}

# expected points per game and game-score standard deviation
moments <- function(x, p, q, th, n) {
  mu <- PTS * p + th
  v  <- PTS^2 * p * (1 - p) + th * (1 - q)
  c(pts = n * sum(x * mu), sd = sqrt(n * sum(x * v)))
}

# feasible set: the widest share any team-season ever demonstrated
S_all <- as.matrix(Z[, SHARE]); S_all <- S_all / rowSums(S_all)
LO <- apply(S_all, 2, min); HI <- apply(S_all, 2, max)
```

The linear program needs the box bounds written as explicit constraint rows, since
`lpSolve::lp()` takes no bounds argument.

```{r lp}
# max  mu'x   s.t.  v'x <= sd^2/n ,  sum x = 1 ,  LO <= x <= HI
frontier_at <- function(sd, p, q, th, n) {
  mu <- PTS * p + th
  v  <- PTS^2 * p * (1 - p) + th * (1 - q)
  M   <- rbind(v, rep(1, 6), diag(6), diag(6))
  dir <- c("<=", "=", rep(">=", 6), rep("<=", 6))
  rhs <- c(sd^2 / n, 1, LO, HI)
  r <- lp("max", mu, M, dir, rhs)
  if (r$status != 0) return(list(pts = NA_real_, x = NULL))
  list(pts = n * r$objval, x = r$solution)
}

# min  v'x   s.t.  mu'x >= m0/n ,  sum x = 1 ,  LO <= x <= HI
min_risk_at <- function(m0, p, q, th, n) {
  mu <- PTS * p + th
  v  <- PTS^2 * p * (1 - p) + th * (1 - q)
  M   <- rbind(mu, rep(1, 6), diag(6), diag(6))
  dir <- c(">=", "=", rep(">=", 6), rep("<=", 6))
  rhs <- c(m0 / n, 1, LO, HI)
  r <- lp("min", v, M, dir, rhs)
  if (r$status != 0) return(NULL)
  list(sd = sqrt(n * sum(r$solution * v)), pts = n * sum(r$solution * mu),
       x = r$solution)
}
```

# The three improvement directions

A holds risk and takes points. B minimises risk subject to not losing points. C is the
point on the dominating frontier segment achieving an equal fraction of each endpoint's
maximum gain.

```{r arbitrage}
arbitrage <- function(season) {
  L <- league(season); th <- theta_of(season)
  m <- moments(L$x, L$p, L$q, th, L$n); m0 <- m[["pts"]]; s0 <- m[["sd"]]

  A <- frontier_at(s0, L$p, L$q, th, L$n)
  B <- min_risk_at(m0, L$p, L$q, th, L$n)
  dmA <- A$pts - m0
  dsB <- s0 - B$sd

  lo <- B$sd; hi <- s0
  for (i in 1:80) {
    mid <- (lo + hi) / 2
    mm <- frontier_at(mid, L$p, L$q, th, L$n)$pts
    if (dsB > 0 && dmA > 0 && (s0 - mid) / dsB > (mm - m0) / dmA) lo <- mid else hi <- mid
  }
  sC <- (lo + hi) / 2; C <- frontier_at(sC, L$p, L$q, th, L$n)

  data.frame(season = season, pts = m0, sd = s0,
             a_dm = dmA, b_ds = dsB, b_dm = B$pts - m0,
             c_dm = C$pts - m0, c_ds = s0 - sC)
}

# Farrell technical efficiency, output- and input-oriented
farrell <- function(season) {
  L <- league(season); th <- theta_of(season)
  m <- moments(L$x, L$p, L$q, th, L$n); m0 <- m[["pts"]]; s0 <- m[["sd"]]
  top <- frontier_at(s0, L$p, L$q, th, L$n)$pts
  smin <- min_risk_at(m0, L$p, L$q, th, L$n)$sd
  data.frame(season = season, eo = m0 / top, ei = smin / s0)
}

ARB <- do.call(rbind, lapply(seasons, arbitrage))
EFF <- do.call(rbind, lapply(seasons, farrell))
SER <- merge(ARB, EFF, by = "season")
```

# Counterfactual paths

Freeze one channel at 2013-14 and let the other move as observed.

```{r counterfactual}
L0 <- league(seasons[1]); th0 <- theta_of(seasons[1])
CF <- do.call(rbind, lapply(seasons, function(s) {
  L <- league(s); th <- theta_of(s)
  sel <- moments(L$x,  L0$p, L0$q, th0, L0$n)   # shot selection moves only
  dev <- moments(L0$x, L$p,  L$q,  th,  L$n)    # shooter development moves only
  act <- moments(L$x,  L$p,  L$q,  th,  L$n)
  data.frame(season = s, mS = sel[["pts"]], sS = sel[["sd"]],
             mD = dev[["pts"]], sD = dev[["sd"]],
             mA = act[["pts"]], sA = act[["sd"]])
}))
SER <- merge(SER, CF, by = "season")
SER$dev_gain <- SER$mD - SER$mD[1]      # points from shooter development
SER$dsd_sel  <- SER$sS - SER$sS[1]      # risk added by shot selection
write.csv(SER, "derived/season_series_recomputed.csv", row.names = FALSE)
knitr::kable(SER[, c("season", "a_dm", "b_ds", "c_dm", "c_ds", "eo", "ei",
                     "dev_gain", "dsd_sel")],
             digits = 3, caption = "Recomputed season series")
```

# Zone tables, attempt-weighted

```{r zone_tables}
zone_table <- function(subset) {
  att <- rep(0, 6); wp <- rep(0, 6); wth <- rep(0, 6); wq <- 0; tq <- 0
  for (s in subset) {
    L <- league(s); th <- theta_of(s)
    nat <- L$x * L$n * games_of(s) * 30            # 30 teams
    att <- att + nat; wp <- wp + nat * L$p; wth <- wth + nat * th
    wq <- wq + sum(nat) * L$q; tq <- tq + sum(nat)
  }
  pb <- wp / att; th <- wth / att; q <- wq / tq
  mu <- PTS * pb + th
  sd <- sqrt(PTS^2 * pb * (1 - pb) + th * (1 - q))
  data.frame(zone = NAME, attempts = att, share = att / sum(att),
             theta = th, exp_pts = mu, sd = sd)
}
TF <- zone_table(seasons)
T26 <- zone_table(max(seasons))
write.csv(TF,  "derived/zone_table_full.csv", row.names = FALSE)
write.csv(T26, "derived/zone_table_2026.csv", row.names = FALSE)
knitr::kable(TF[order(-TF$attempts), ], digits = 3, row.names = FALSE,
             caption = "2013-14 through 2025-26 shot selection data summary")
knitr::kable(T26[order(-T26$attempts), ], digits = 3, row.names = FALSE,
             caption = "2025-26 shot selection data summary")
```

# Zone-share regressions

```{r regressions}
zone_reg <- function(T) {
  d <- data.frame(share = T$share, ExpPts = T$exp_pts, SD = T$sd)
  fits <- list(both = lm(share ~ ExpPts + SD, d),
               ret  = lm(share ~ ExpPts, d),
               risk = lm(share ~ SD, d))
  lapply(fits, function(f) list(coef = round(coef(f), 3),
                                r2 = round(summary(f)$r.squared, 3)))
}
RF <- zone_reg(TF); R26 <- zone_reg(T26)
for (tag in c("full", "2026")) {
  r <- if (tag == "full") RF else R26
  cat(sprintf("%s  zone-share = %+.3f %+.3f*ExpPts %+.3f*SD   R2 = %.3f\n",
              tag, r$both$coef[1], r$both$coef[2], r$both$coef[3], r$both$r2))
  cat(sprintf("%s  return alone R2 = %.3f   risk alone R2 = %.3f\n",
              tag, r$ret$r2, r$risk$r2))
}
```

# Verification against the abstract

Every quantity printed in the abstract, recomputed here and compared. The chunk stops
the knit if any check fails.

```{r verify}
r26 <- SER[SER$season == max(seasons), ]
checks <- list(
  list("A: points at unchanged risk",       r26$a_dm,              6.68,  0.02),
  list("B: risk shed",                      r26$b_ds,              1.22,  0.02),
  list("C: points",                         r26$c_dm,              5.68,  0.03),
  list("C: risk",                           r26$c_ds,              1.04,  0.03),
  list("Farrell output-oriented",           r26$eo,                0.945, 0.002),
  list("Farrell input-oriented",            r26$ei,                0.892, 0.002),
  list("shooter-development gain",          tail(SER$dev_gain, 1), 9.32,  0.05),
  list("SD added by shot selection",        tail(SER$dsd_sel, 1),  0.70,  0.03),
  list("vertical shortfall 2013-14",        SER$a_dm[1],           4.08,  0.03),
  list("total FGA, millions",               sum(TF$attempts)/1e6,  2.776, 0.02),
  list("rim expected points 2025-26",       T26$exp_pts[1],        1.662, 0.01),
  list("rim SD 2025-26",                    T26$sd[1],             0.941, 0.01),
  list("above-break 3 exp. pts 2025-26",    T26$exp_pts[6],        1.175, 0.01),
  list("above-break 3 share 2025-26 (%)",   100*T26$share[6],      31.3,  0.3),
  list("rim share 2025-26 (%)",             100*T26$share[1],      23.1,  0.3),
  list("R2 zone-share full, both",          RF$both$r2,            0.273, 0.005),
  list("R2 zone-share full, risk only",     RF$risk$r2,            0.001, 0.005),
  list("R2 zone-share 2025-26, both",       R26$both$r2,           0.245, 0.005),
  list("R2 zone-share 2025-26, risk only",  R26$risk$r2,           0.059, 0.005)
)
res <- do.call(rbind, lapply(checks, function(k) data.frame(
  quantity = k[[1]], recomputed = round(as.numeric(k[[2]]), 3),
  abstract = k[[3]],
  status = ifelse(abs(as.numeric(k[[2]]) - k[[3]]) <= k[[4]], "PASS", "FAIL"))))
knitr::kable(res, row.names = FALSE, caption = "Reproduction check")
cat(sprintf("\n%d of %d checks pass\n", sum(res$status == "PASS"), nrow(res)))
if (any(res$status == "FAIL")) stop("reproduction check failed; see the table above")
```

# Figure 1 — two counterfactuals

The surface is the frontier, the line the actual path. Left panel freezes shooter
development at 2013-14 with shot selection as observed; right panel freezes shot
selection with development as observed.

`ggplot2` has no 3D geometry, so the default below is a faceted 2D rendering of the same
two surfaces: a filled raster of the frontier with the actual path drawn over it. This
reads the gradient more precisely than a perspective view and has no occlusion. A
`plot3D::persp3D` version follows for the perspective look.

```{r fig1_surfaces}
risk_grid <- seq(10.1, 12.1, length.out = 26)
surf <- do.call(rbind, lapply(seasons, function(s) {
  L <- league(s); th <- theta_of(s)
  data.frame(season = s, risk = risk_grid,
             dev_frozen = sapply(risk_grid, function(v)
               frontier_at(v, L0$p, L0$q, th0, L0$n)$pts),
             sel_frozen = sapply(risk_grid, function(v)
               frontier_at(v, L$p, L$q, th, L$n)$pts))
}))
write.csv(surf, "derived/cf_surface_grid.csv", row.names = FALSE)
```

```{r fig1_2d, fig.width=11, fig.height=4.4}
long <- surf %>%
  pivot_longer(c(dev_frozen, sel_frozen), names_to = "panel", values_to = "frontier") %>%
  mutate(panel = factor(panel, c("dev_frozen", "sel_frozen"),
           c("Shooter development frozen, shot selection as observed",
             "Shot selection frozen, shooter development as observed")))
paths <- bind_rows(
  transmute(CF, season, risk = sS, pts = mS,
            panel = "Shooter development frozen, shot selection as observed"),
  transmute(CF, season, risk = sD, pts = mD,
            panel = "Shot selection frozen, shooter development as observed"))

ggplot(long, aes(season, risk)) +
  geom_raster(aes(fill = frontier), interpolate = TRUE) +
  geom_contour(aes(z = frontier), colour = "white", linewidth = 0.25, bins = 12) +
  geom_path(data = paths, colour = "white", linewidth = 2.1) +
  geom_path(data = paths, colour = "grey12", linewidth = 1.1) +
  geom_point(data = paths, colour = "white", fill = "grey12",
             shape = 21, size = 1.7, stroke = 0.6) +
  facet_wrap(~ panel) +
  scale_fill_viridis_c(option = "inferno", name = "frontier\n(points)") +
  labs(x = NULL, y = "risk: game-score standard deviation",
       title = "Figure 1. The frontier and the actual path under each counterfactual",
       subtitle = paste("Left: the frontier is flat in season and the path slides right.",
                        "Right: the frontier rises and the path rises with it.")) +
  theme_minimal(base_size = 11) +
  theme(panel.grid = element_blank(), strip.text = element_text(face = "bold"),
        plot.title = element_text(face = "bold"), plot.title.position = "plot")
ggsave("figures/figure1_counterfactuals.pdf", width = 11, height = 4.4)
```

```{r fig1_3d, eval=FALSE}
# perspective version; set eval=TRUE and install plot3D to use it
library(plot3D)
yrs <- seasons; rsk <- risk_grid
ZA <- matrix(surf$dev_frozen, nrow = length(yrs), byrow = TRUE)
ZB <- matrix(surf$sel_frozen, nrow = length(yrs), byrow = TRUE)
panel <- function(M, px, py, pz, pal, ttl) {
  persp3D(x = yrs, y = rsk, z = M, col = pal(100), colkey = FALSE,
          theta = -62, phi = 20, expand = 0.6, border = NA, shade = 0.25,
          alpha = 0.55, bty = "b2", xlab = "season",
          ylab = "risk: scoring SD", zlab = "points per game",
          main = ttl, cex.main = 0.95)
  lines3D(px, py, pz, col = "white", lwd = 6, add = TRUE)
  lines3D(px, py, pz, col = pal(100)[85], lwd = 4, add = TRUE)
}
par(mfrow = c(1, 2), mar = c(2, 2, 3, 1))
panel(ZA, CF$season, CF$sS, CF$mS,
      colorRampPalette(c("#FFF7BC", "#C0392B")),
      "(a) Shooter development frozen")
panel(ZB, CF$season, CF$sD, CF$mD,
      colorRampPalette(c("#F7FCB9", "#1B7A4B")),
      "(b) Shot selection frozen")
```

# Figure 2 — the frontier and where attempts should go

```{r fig2, fig.width=11, fig.height=4.6}
sn <- max(seasons); L <- league(sn); th <- theta_of(sn)
m <- moments(L$x, L$p, L$q, th, L$n); m0 <- m[["pts"]]; s0 <- m[["sd"]]
B <- min_risk_at(m0, L$p, L$q, th, L$n)
A <- frontier_at(s0, L$p, L$q, th, L$n)
fr <- data.frame(risk = seq(B$sd, 12.0, length.out = 80))
fr$pts <- sapply(fr$risk, function(v) frontier_at(v, L$p, L$q, th, L$n)$pts)

# the league's thirteen-season path
trail <- data.frame(season = CF$season, risk = CF$sA, pts = CF$mA)
# 2013-14 frontier, for the ghost line
fr0 <- data.frame(risk = fr$risk,
                  pts = sapply(fr$risk, function(v)
                    frontier_at(v, L0$p, L0$q, th0, L0$n)$pts))

p_left <- ggplot() +
  geom_ribbon(data = subset(fr, risk <= s0 & pts >= m0),
              aes(risk, ymin = m0, ymax = pts), fill = "#C0392B", alpha = 0.09) +
  geom_line(data = fr0, aes(risk, pts), colour = "#BFCAD1", linewidth = 1.1) +
  geom_line(data = fr,  aes(risk, pts), colour = "#14222B", linewidth = 1.3) +
  geom_path(data = trail, aes(risk, pts), colour = "#E0A79E", linewidth = 0.7) +
  geom_point(data = trail, aes(risk, pts, colour = season), size = 1.8) +
  scale_colour_gradient(low = "#F3D3CD", high = "#C0392B", guide = "none") +
  annotate("point", x = s0, y = m0, shape = 8, size = 4, colour = "#C0392B") +
  annotate("segment", x = s0, xend = s0, y = m0, yend = A$pts,
           colour = "#1B7A4B", linewidth = 1.0,
           arrow = arrow(length = unit(0.16, "cm"))) +
  annotate("segment", x = s0, xend = B$sd, y = m0, yend = B$pts,
           colour = "#1F6FB2", linewidth = 1.0,
           arrow = arrow(length = unit(0.16, "cm"))) +
  labs(x = "risk: game-score standard deviation", y = "return: points per game",
       title = "Thirteen seasons inside the frontier") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"), panel.grid.minor = element_blank())

redist <- data.frame(zone = NAME, league = L$x / sum(L$x), target = B$x) %>%
  mutate(zone = factor(zone, levels = NAME[order(L$x)]))
p_right <- ggplot(redist, aes(y = zone)) +
  geom_segment(aes(x = league, xend = target, yend = zone),
               arrow = arrow(length = unit(0.16, "cm")),
               colour = "#1B4965", linewidth = 0.9) +
  geom_point(aes(x = league), colour = "#C0392B", size = 2.6) +
  geom_text(aes(x = 0.47, label = sprintf("%+.3f", target - league)),
            hjust = 1, size = 3) +
  xlim(0, 0.48) +
  labs(x = "share of field goal attempts", y = NULL,
       title = "How 2025-26 should redistribute") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"), panel.grid.minor = element_blank())

if (requireNamespace("patchwork", quietly = TRUE)) {
  print(patchwork::wrap_plots(p_left, p_right, widths = c(1.25, 1)))
} else {
  print(p_left); print(p_right)
}
ggsave("figures/figure2_frontier.pdf", plot = p_left, width = 6.2, height = 4.4)
ggsave("figures/figure2_redistribution.pdf", plot = p_right, width = 5.0, height = 4.4)
```

# Three-point volume trend

```{r trend, fig.width=8, fig.height=3.6}
TR <- do.call(rbind, lapply(seasons, function(s) {
  L <- league(s)
  data.frame(season = s, w = sum(L$x[5:6]), TPA = L$n * sum(L$x[5:6]), FGA = L$n)
}))
ggplot(TR, aes(season)) +
  geom_line(aes(y = w), colour = "#C0392B", linewidth = 1.1) +
  geom_point(aes(y = w), colour = "#C0392B", size = 1.6) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(x = NULL, y = "three-point attempt rate",
       title = "Three-point attempt rate, 2013-14 to 2025-26") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"), panel.grid.minor = element_blank())
ggsave("figures/trend_3par.pdf", width = 8, height = 3.6)
```

# Session information

```{r session}
sessionInfo()
```
