---
title: "Three-Point Volume Over the Tracking Era"
subtitle: "Figure source for *Modern NBA Shot Selection: More Risk for Less Return in a First-Order Analytic League*"
author: "Shane Sanders"
output:
  pdf_document:
    keep_tex: true
    fig_width: 10.6
    fig_height: 4.3
geometry: margin=0.9in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      dev = "pdf", fig.align = "center")
library(ggplot2)
library(dplyr)
library(patchwork)   # for the side-by-side layout
```

## Data

League averages, 2013-14 through 2025-26, from the team-season panel. The tracking era begins with
league-wide SportVU installation in 2013-14.

```{r data}
S <- read.csv("trend_series.csv")          # season, yr, w (3PAr), TPA, FGA
S <- S %>% mutate(yr = as.integer(yr))
growth <- function(x) 100 * (tail(x, 1) / head(x, 1) - 1)
g_tpa <- growth(S$TPA); g_w <- growth(S$w)
c(TPA = g_tpa, PAr = g_w)
```

## Theme

A stripped-back theme: no panel border, no vertical grid, no tick marks, muted axis text, and a
left-aligned title with a subtitle carrying the takeaway.

```{r theme}
INK  <- "#12232E"; LINE <- "#1B4965"; DOT <- "#D1495B"; MUTE <- "#8FA6B2"

theme_court <- function(base_size = 11) {
  theme_minimal(base_size = base_size) +
    theme(
      panel.grid.minor   = element_blank(),
      panel.grid.major.x = element_blank(),
      panel.grid.major.y = element_line(colour = "#E6EBEE", linewidth = 0.35),
      panel.border       = element_blank(),
      axis.ticks         = element_blank(),
      axis.text          = element_text(colour = MUTE, size = rel(0.92)),
      axis.title         = element_text(colour = "#4A5C66", size = rel(0.88)),
      axis.title.x       = element_blank(),
      plot.title         = element_text(colour = INK, face = "bold",
                                        size = rel(1.30), hjust = 0),
      plot.subtitle      = element_text(colour = "#6B7C86", size = rel(0.90), hjust = 0),
      plot.title.position = "plot",
      plot.caption       = element_text(colour = "#8A98A1", size = rel(0.78), hjust = 0.5),
      plot.margin        = margin(10, 14, 6, 10)
    )
}
```

## Figure

Each panel fills lightly under the line, marks every season with an open point, and labels only
the endpoints in the accent colour. Direct labelling replaces a legend entirely.

```{r figure, fig.width=10.6, fig.height=4.3}
panel <- function(df, yvar, ylab, ttl, sub, digits, ylim, brks) {
  ends <- df[c(1, nrow(df)), ]
  ggplot(df, aes(x = yr, y = .data[[yvar]])) +
    geom_area(fill = LINE, alpha = 0.07) +
    geom_line(colour = LINE, linewidth = 1.05, lineend = "round") +
    geom_point(shape = 21, fill = "white", colour = LINE,
               size = 1.9, stroke = 0.75) +
    geom_point(data = ends, colour = DOT, size = 2.6) +
    geom_text(data = ends,
              aes(label = formatC(.data[[yvar]], format = "f", digits = digits)),
              colour = DOT, fontface = "bold", size = 4.0,
              hjust = c(1.25, -0.25), vjust = c(1.1, -0.6)) +
    scale_x_continuous(breaks = seq(2014, 2026, 2),
                       labels = function(v) paste0("'", substr(v, 3, 4)),
                       expand = expansion(mult = 0.06)) +
    scale_y_continuous(breaks = brks) +
    coord_cartesian(ylim = ylim, clip = "off") +
    labs(title = ttl, subtitle = sub, y = ylab) +
    theme_court()
}

p1 <- panel(S, "TPA", "attempts per team-game",
            "Three-point attempts",
            sprintf("per team-game, up %.0f%% since tracking began", g_tpa),
            1, c(20.5, 39.5), seq(22, 38, 4))

p2 <- panel(S, "w", "share of field goal attempts",
            "Three-point attempt rate",
            sprintf("3PAr, up %.0f%% since tracking began", g_w),
            3, c(0.243, 0.437), seq(0.26, 0.42, 0.04))

(p1 | p2) +
  plot_annotation(
    caption = paste("NBA league averages, 2013-14 through 2025-26.",
                    "The tracking era begins with league-wide SportVU installation in 2013-14."),
    theme = theme_court()
  )
```

```{r save, eval=FALSE}
ggsave("trend3par.pdf", (p1 | p2), width = 10.6, height = 4.3, device = cairo_pdf)
```

## Note on the two panels

The curves are nearly identical in shape, which is worth stating in the text: the rise in
three-point volume was not a pace artifact. Field goal attempts per game moved only from
`r sprintf("%.1f", S$FGA[1])` to `r sprintf("%.1f", tail(S$FGA, 1))`, so a `r sprintf("%.0f", g_tpa)`
percent rise in attempts and a `r sprintf("%.0f", g_w)` percent rise in the rate are the same
event. The league reallocated a roughly fixed budget of possessions.
